---
title: "Exploratory Analysis: Loan Data"
graphics: yes
author: "Nikhil Haas"
date: "August 10, 2015"
output: pdf_document
header-includes:
    - \usepackage{setspace}
    - \doublespacing
    - \usepackage{float}
---

```{r setup, include=FALSE}
options(warn=-1); suppressMessages(library(knitr)); options(warn=0)
knitr::opts_chunk$set(cache=2)
source("geo_plots.R")
source("stats_plots.R")
source("hist_plots.R")
load_package("png")
loans <- read_loans_data()
```

# Introduction

A data set containing 113,937 rows of data and 81 variables was downloaded. Each
row contains information on a loan, including loan amount, borrower rate (or
interest rate), current loan status, borrower income, borrower employment
status, borrower credit history, and the latest payment information. Data is
from Prosper.com, a peer-to-peer lending marketplace. Prosper.com allows
individuals to invest in other individuals and handles the loan servicing on
behalf of the borrower and investor. Let's ask a couple questions or form a
hypothesis about the relationships between some of the variables. These
questions will be explored with several plots and analysis.

# Questions

Several questions come to mind, considering that this data set could potentially
contain information that 1) could illustrate macroeconomic trends, 2)
demonstrate how the loan industry or investors might use various criteria to
approve loans, and 3) show whether or not those criteria indeed give loan
originators enough information make a prediction about the risk of loaning to
a borrower.

  * Macroeconomic relationships
      1. Can loan origination by state be a proxy for economic health?
      2. Is there a trend in delinquency rate coinciding with a recession?
      3. Are individuals with certain job titles more or less likely to apply
         for a loan?
  * Inter-variable relationships
      4. What relationships (linear or otherwise) exist in the data?
      5. Do people with certain occupations have more income
         or delinquencies, or are they able borrow more than others?
      6. Is credit score a good indicator of on-time payments?
      7. Is number of past delinquincies an indicator of future delinquincies?

# Macroeconomics Relationships

## Geographic

We can begin by visualizing each loan origination by state. There were 871
duplicate values for _LoanKey_, which were removed prior to aggregating loans by
state. Each _LoanKey_ is a unique identifier for a given loan, so the duplicate
entries might contain information about a loan from a different point in time.

```{r plot_originations_by_state, message=FALSE, echo=FALSE, fig.width=6.21, fig.height=3.20, fig.align='center', fig.pos='b', fig.cap="Loan originations by state."}
plot_loans_by_state()
```

```{r plot_gdp_by_state, message=FALSE, echo=FALSE, fig.width=6.83, fig.height=3.41, fig.align='center', fig.cap="US states by GDP, 2008. Source: Wikipedia.org."}
load_package("png")
library(grid)
img <- readPNG("us-gdp-states-2008.png")
grid.raster(img)
```

In Fig. 1, we see that a large portion of loan borrower applications are from
California. Texas, Florida, New York, and Illinois also stand out on the map.
From a business standpoint, this might or might not give us much information,
depending on how the company is marketing and where its user-base is.
Interestingly, we see some parallels between the US states by GDP (Fig. 2) and
the states from which many borrowers originate: the top states by GDP are also
the top borrower states. These states are also likely the most populated, as
well, which could be the reason why more loans come from those states. This
kind of analysis gives us an over-arching view of the geospacial information
contained in the data set that could help us to dive in deeper, much like
a histogram gives us an overview of data before moving on to detailed
statistical analysis. Further geospacial studies could help the business
identify certain market opportunities.

We can also plot trends on the US map. The data contains loan origination dates,
and by separating the data by year and calculating the percent difference in
loans YoY, we have a new dimension (time) to incorporate. In Fig. 3 we can see
which states are experiencing the most (and least) growth in loan originations
over time. This might tell us about the sales history and trajectory of
Propser.com and could be relevant for any marketing groups in the company.
Fig. 3 could also be compared with trends in GDP by state to see if any states
experienced negative growth when compared with change in GDP, which might
indicate that Propser.com is losing market share in that state. We observe
that no state experienced negative growth, so this should indicate that the
business is growing overall, with some states bringing in more business at a
faster rate than others.

```{r plot_trends_by_state, message=FALSE, echo=FALSE, fig.width=6.52, fig.height=3.10, fig.align='center', fig.pos='H', fig.cap="Trends in originations by state. Gray indicates no data available."}
plot_state_trends()
```

A lot of other interesting information was garnered from similar state maps,
such as which states experienced the greatest default rates (not included).
Geographic data could be explored _ad infinitum_ - every statistical measure
we make could be broken down at the state level - but let's move on so we can
explore other aspects of the data.

## Occupation

There are approximately 63 occupations listed in the loans data. The occupations
were consolidated into 43 occupations. For example, "Student - College" and
"Student - High School" were consolidated into a single occupation "Student",
In Fig. 4 we can see the loan originations by occupation. Certain occupations
appear more than others. This doesn't necessarily mean that certain
occupations are more likely to borrow or get approved to borrow; this is more
likely a representation of the user base of Prosper.com.

The category of _Professional_ is almost three times larger than most of the
other categories. If the company were to try to use occupation to make
predictions about or to better understand borrowers, Fig. 4 might suggest
splitting up the _Professional_ category on loan applications. This could
potentially allow some hidden sub-categories of "Professional" occupations to
surface. It could be the case that _Professional_ has so many varied meanings
that the category is useless as a predictor. Imagine if a third of Professionals
identified themselves as "Technical Professionals" and another third identified
as "Administrative Professionals". It could be the case that "Technical
Professional" always defaulted on their loans and "Administrative
Professional" never did, but since all those individuals are grouped into
"Professional" that trend would never be exposed.

```{r plot_occupation, message=FALSE, echo=FALSE, fig.width=7, fig.height=4, fig.align='center', fig.cap="Loan data by Occupation", fig.pos='H'}
plot_occupation_histogram()
```

We might wonder if certain occupations experience more delinquencies than
others, or if certain occupations are more likely to complete their loan
payments. By calculating the frequencies of _LoanStatus_ by occupation (Fig. 5),
we can see that there are a high number of completed loans for students and a
low number of completed loans for judges. This might seem unusual if at first
glance one
assumes that this indicates students are more reliable borrowers and judges are not
(which would contradict our presumed understanding of those two populations).
However, it is important to realize that these observed trends do not
necessarilyy indicate causation. It is possible that Prosper.com first broke into
the Student market and therefore that category of occupation shows a high
percentage of completed loans; judges, on other hand, might only be recently
exposed to Propser.com as a means to borrow money (e.g. now that Prosper.com is
growing in popularity). If several judges recently opened loans, they might
skew the proportion of loans that are current.

```{r plot_status_by_occupation, message=FALSE, echo=FALSE, fig.width=7.5, fig.height=6.1, fig.align='center', fig.pos='H', fig.cap="Loan status by occupation"}
plot_status_by_occupation()
```

# Inter-variable Relationships

## Relationships to Credit Score

We can pull out the numeric columns from the data and compare each column
against others. Creditors often use credit score as one of the primary measures
by which a borrower is rated, so let's explore relationships between various
measure about a loan and the borrower's credit score. Here, we use the upper
value of the borrower's credit score as reported by a credit agency,
_CreditScoreRangeUpper_. As a creditor, we would be interested in exploring
whether or not credit score is correlated with delinquencies. As
observers of the loan industry, we can compare measures like
_DelinqueniciesLast7Years_, _RevolvingCreditBalance_, _DebtToIncomeRatio_
to see how a borrower's financial picture influences his or her credit score.
As a borrower, we might be interested in seeing how our credit score changes
the _BorrowerRate_. Furthermore, we can see whether or not there is a strong
link between credit score and _ProsperScore_, the rating Prosper.com assigns
to the borrower. In effect, we want to see what (if any) relationships exist
in the data between credit score and other measures. The measures
are scaled between 0 and 1 because we are comparing credit score with measures
that have very different ranges - some are fractions representing a percent,
others; such as monthly income, are orders of magnitude larger. We scale so we
can plot on one axis.

In Fig. 6, which plots the relationships between credit score and these
various measures, we can begin to understand some of these relationships. The
discussion of these relationships will be short since the graph is fairly
self-explanatory.

  * **Credit score and delinquencies:** delinquencies drop at the upper and
  lower ranges of credit scores. Fewer delinquencies in the last seven years is
  correlated with higher credit scores.
  * **Credit available and usage:** higher _AvailableBankcardCredit_ appears to
  be strongly correlated with high credit scores, and _CurrentDelinquencies_
  inversely correlated with high credit scores. 
  * **Monthly income:** there do not appear to be any significant correlations
  between monthly income and credit score.
  * **Borrower rate:** _BorrowerRate_ looks be be capped for credit scores below
  approximately 550. Borrower rates are fairly evenly distributed across credit
  scores, which might indicate that there are other factors with significant
  influcene on _BorrowerRate_ (such as _DebtToIncomeRatio_), or that
  _BorrowerRate_ does not depend on credit score.
  * **Prosper score:** scores assigned to a loan by Prosper seem to be evenly
  distributed, except very high credit scores generally receive a high
  _ProsperScore_.
  
```{r save_credit_score_plot, include=FALSE, message=FALSE, echo=FALSE, fig.height=8.5}
# Don't put plot in PDF directly - it plots each individual point into the PDF
plot_against_credit_score()
```

```{r insert_credit_score_plot, message=FALSE, echo=FALSE, fig.align='center', fig.height=8.5, fig.pos='H', fig.cap="Scatter of various measurements by credit score with density in the background"}
# Here we insert the PNG created by save_credit_score() above
load_package("png")
library(grid)
img <- readPNG("compare-credit-score.png")
grid.raster(img)
```
  
This type of plot (a facet wrap) lets us easily compare multiple variables
and multiple categories. We could increase the complexity, for example
by making point size scale with the amount borrowed, to visualize even more
data. However, the point of this graph is to identify correlations so we keep
additional data off this plot to make it easier to identify patterns as credit
score increases or decreases. Since the data is shown as a scatter plot,
densities of the scatter have been placed behind the points so that we
can recognize when trends in the scatter are due to variability in the data
rather than a trend. We might also consider taking the second quantile (Q2) of
data or removing outliers prior to plotting so they do not distract our eyes
from possible trends.

# Conclusion

There are a lot of plots and endless analysis that can be done on this data.
With so many columns and a decent number of rows there are many visualizations
that could be created with a specific hypothesis in mind. In this short report,
I sought to explore general economic trends, get a basic idea of how
credit score is impacted by a borrower's financial history, and see how credit
score might influence his or her loans with Prosper.com.
